---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Peer Assessment 1

``` {r, echo = TRUE}
 ## loading required R packages
  require(dplyr)
  require(xtable)
  require(ggplot2)
```


## Loading and preprocessing the data

### Load the data
-Load the data from file activity.csv located in working directory  
-if data does not exist in global environment  

``` {r, echo = TRUE}
## check if dataframe exists
  dfExists <- function(df_name){
    
    return(exists(df_name) && is.data.frame(get(df_name)))
    
  }  

## Load data if data does not exist in global environment
  
  if(dfExists('actdata') == FALSE ){
    
    actdata <<- read.csv("activity.csv")
    }
```
### Process the data

``` {r, echo = TRUE}
  ## rearrange data by date
  actdata <<- arrange(actdata,date)
  
  ## group data by date
  byday <<- group_by(actdata,date)
  
  ## group data by interval
  byinterval <<- group_by(actdata,interval)
```

## What is mean total number of steps taken per day?

### Histogram of the total number of steps taken each day
``` {r hist, fig.height = 5}
 
  bydaysteps <- summarize(byday,sum_steps=sum(steps))
  
  hist(x = bydaysteps$sum_steps, main = "Total number of steps taken per day", xlab = "Steps" )

```

### Mean total number of steps taken each day
``` {r showmean, results="asis"}

  xt <- xtable(summarize(byday,mean_steps=mean(steps,na.rm=TRUE)))
 
  print(xt, type = "html")

```
### Median total number of steps taken per day
``` {r showmedian, results="asis"}
 xt <- xtable(summarize(byday,median_steps=median(steps,na.rm=TRUE)))

  print(xt, type = "html")

```
 
## What is the average daily activity pattern?

```{r timeseries, fig.height = 5}

maketimeseriesplot <- function(){
  
  plotdata <- summarize(byinterval,mean_steps = mean(steps, na.rm=TRUE))
  
  theplot <- ggplot(plotdata, aes(x=interval,y=mean_steps))
  theplot <- theplot + geom_line()
  theplot <- theplot + xlab("Interval")
  theplot <- theplot + ylab("Number of Steps")
  theplot <- theplot + ggtitle("5-minute interval and average number of steps taken across all days")
  
  theplot
  
}

maketimeseriesplot()

```
### Which 5 minute interval, on average across all days contains the maximum number of steps?

```{r}
calcmaxstepinterval <- function(){
  
  plotdata <- summarize(byinterval,mean_steps = mean(steps, na.rm=TRUE))
  
 maxsteprow <- plotdata[plotdata$mean_steps == max(plotdata$mean_steps), ]
 
 return(maxsteprow$interval)
  
  
}

```
### The 5 minute interval is `r calcmaxstepinterval()`.  

## Imputing missing values

### What is the total number of missing values in the dataset?
```{r}

## Calculate total number of missing values 

calctotalmissingvalues <- function(){
  
 numberofrows = nrow(actdata)
 
 numberofnonNArows = nrow(na.omit(actdata))
 
 return(numberofrows-numberofnonNArows)
 
}

``` 

### The total number of missing values is `r calctotalmissingvalues()`.


## Strategy for filling in all of the missing values in the dataset
### If value for steps is NA then the mean value for the 5 minute interval is substituted for NA.
### The new data are stored in a dataframe named imputeddata.

```{r}

## Create dataset that is equal to the original dataset but with the missing data filled in.

createimputeddata <- function(){
  
  intervalmeans <- summarize(byinterval,mean_steps_by_interval=mean(steps,na.rm=TRUE))
  imputeddata <<- merge(byday,intervalmeans,by=c("interval"))
  imputedbyday <<- arrange(imputeddata, date)
  imputedbyday <<- group_by(imputedbyday,date)
  imputeddata <<- transform(imputedbyday, steps = ifelse(is.na(steps)==FALSE,steps,mean_steps_by_interval))
  imputeddata <<- group_by(imputeddata,date)
  imputeddata <<- arrange(imputeddata,date) 
}

createimputeddata()

```

### Total number of steps taken each day based upon the imputed data set
```{r hist2, fig.height = 5}
## Create histogram of total number of steps taken each day for imputed dataset
makehist_imputeddata <- function(){
  
  bydaysteps <- summarize(imputeddata,sum_steps=sum(steps))
  
  hist(x = bydaysteps$sum_steps, main = "Total number of steps taken per day 
       \n if steps = NA then use interval mean", xlab = "Steps" )
  
}

makehist_imputeddata()
```

## Calculate and report the mean and median total number of steps taken per day.
``` {r showimputedmean, results = "asis"}

require(ggplot2)

## Calculate mean number of steps taken per day for imputed data

  xt <- xtable(summarize(imputeddata,mean_steps_imputed=mean(steps,na.rm=TRUE)))
  
  print(xt, type = "html")

```
```{r showimputedmedian, results = "asis"}
## Calculate median number of steps taken per day for imputed data

  xt <- xtable(summarize(imputeddata,median_steps_imputed=median(steps,na.rm=TRUE)))
  
  print(xt, type = "html")

```

## Are there differences in activity patterns between weekdays and weekends?

```{r panel, fig.height = 5}

## Create new factor variable indicating wheter given date is a weekday or weekend

weekdayorweekend <- function(thisdate){
  
  thedate = as.Date(thisdate)
  
  weekdayname = weekdays(thedate)
  
  if(weekdayname == "Sunday" || weekdayname == "Saturday"){
    return("weekend")
  }
  else
  {
    return("weekday")
    
  }
}

addweekdayfactor <- function(){
  
  dayfactordataset <<- mutate(imputeddata,dayfactor = weekdayorweekend(date))
  
  dayfactordataset <<- group_by(dayfactordataset,dayfactor,interval)
   
}

## Make a panel plot containing time series plot of the 5 minute interval and average number of steps taken 
## averaged across all weekday days or weekend days

makepanelplot <- function(){
  
  plotdata <- summarize(dayfactordataset,mean_steps = mean(steps))
  
  theplot <- ggplot(plotdata, aes(x=interval, y=mean_steps))
  theplot <- theplot + geom_line()
  theplot <- theplot + facet_wrap(~ dayfactor)
  theplot <- theplot + facet_grid(dayfactor ~ .) 
  theplot <- theplot + xlab("Interval")
  theplot <- theplot + ylab("Number of Steps")
  theplot <- theplot + ggtitle("Activity patterns between weekdays and weekends")
  
  theplot
  
}

addweekdayfactor()

makepanelplot()
```

